---
title: "STA 207: Assignment I"
author: "Elsie Basa Student ID:914022828"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(SuppDists)
library(gplots)
library(MASS)
```

```{r, echo = FALSE}

Wage=read.csv('Wage.csv')

attach(Wage, warn.conflicts = F)
```

###1.
The one way ANOVA model using the f actor-effect form is:*
$$ Y_{ij} = \mu + \alpha_i + \epsilon_{ij} \quad for \space i = 1,..,6 \space and \space j=1,..,n$$

*where $\epsilon_i$ are i.i.d $N(0,\sigma^2)$ and $\{ \alpha_i\}$ satisfies that $\sum_{i=1}^6 n_i \alpha_i=0$*


###2.
*The least square estimator for $\alpha_i$ is $\hat\alpha_i$ =  $\hat\mu_i$ - $\hat\mu$ = $\bar{Y_{i.}}$ - $\bar{Y_{..}}$ *

*The Expectation*

$$E(\hat\alpha_i) = E(\hat\mu_i - \hat\mu) = \hat\mu_i - \hat\mu = \alpha_i $$

The expectation in terms of $n_i$:
$$ E[\hat\alpha_i]=E\left[\left(\bar{Y}_{i .}-\bar{Y}_{. .}\right)^2\right]$$

$$=\left[\left(\frac{1}{n_i}-\frac{1}{N}\right) \sigma^2+\left(\mu_i-\mu\right)^2\right]$$

*The Variance:* 
$$ Var(\hat\alpha_i) = Var(\hat\mu_i - \hat\mu) = Var(\bar{Y_{i.}} -\bar{Y_{..}})  $$


$$ = Var \left( \bar{Y_{i.}} -\frac {\sum_{s=1}^k n_s\bar{Y}_{s .}}{N} \right) $$


$$ = Var \left(\left(1-\frac{n_i}{N}\right) \bar{Y}_{i .}-\frac{\sum_{s \neq i} n_s \bar{Y}_{s .}}{N}\right) $$

$$=\left(1-\frac{n_i}{N}\right)^2  Var\left(\bar{Y}_{i .}\right)+\sum_{s \neq i} \frac{n_s^2}{N^2} \operatorname{Var}\left(\bar{Y}_{s .}\right) $$

$$=\left(1-\frac{n_i}{N}\right)^2 \frac{\sigma^2}{n_i}+\sum_{s \neq i} \frac{n_s^2}{N^2} \frac{\sigma^2}{n_s}$$

$$ = (\frac{1}{n_i} - \frac{1}{N})\sigma^2 $$




*3.*
*Main effect Plot:* 

*From the Occupaction main effect plot it shows that management occupation earns the highest mean wage, followed by a technical occupation. While the lowest average wage are for those with a service occupation. Managment has the most variability*



```{r, echo = FALSE}

# Main effect plot for occupation
plotmeans(wage~occupation,data=Wage, xlab ="Occupation",ylab="Wage",
          main="Main  effect, Occupation ",cex.lab=1.2) 


```


###4.

*ANOVA Table*
```{r, echo = FALSE}
anova_fit<- aov(wage~ (occupation),data=Wage)

#summary of aov()
summary <- summary(anova_fit) ; summary
```

*This ANOVA Table shows that the degrees of freedom which is the number of indpendent values that can vary in our model groups. The degrees of freedom for occupation is 6 - 1 which is 5. The degrees of freedom for the residuals are 534, the number of observations, minus 6, the number of occupation groups.  The table also include sum of squares which represent the squared devations from the mean which are then summed up to get the sum of squared values. Next, in the table is the Mean sum of squares which are the sum of squared values divided by the degrees of freedom. Next, the F-Value in the table is a ratio between the mean squared terms. This value can be used to test if the means amoung each of the groups are equal. Finally, the pvalue is used to determine if the statistical significance of the F test. *




###5.

####a)

$$ H_o: \text{All} \space \mu_i \space \text{are the same from} \space\text{i = 1 to 6} \quad vs \quad H_a:\space \text{not all} \space \mu_i \space are \space 0
 $$




####b)
*To test if there is an interaction between wages and occupation an F test at a 0.05 signficance level. The distribution of the F-Stat is an F distribution.  To calculate* $F^*$, $F^*$ = MSTR/MSE =  507.5/ 21.9 *which is 23.22392. The F critcal value of the test is 2.231087.* *So,* $F^*$ > *F critical value* 

*Note:MSTR is Mean Squared Treatment and MSE is Mean Square Error
 

####c) 
*In the test* $F^*$ = 23.22392  *which is greater than the $Fcrit$ of 2.231087. Therefore we reject the null hypothesis and therefore there is signficant evidece to suggest that there is an associationn between wages and occupation.*



```{r, echo = FALSE, include=FALSE }
r=anova_fit$rank ;n=length(Wage$occupation)
alpha=0.05
#x.grid=seq(from=1e-5,to=6,length.out=1000);

crit.val=qf(1-alpha,df1=r-1,df2=n-r);
fstar = summary(anova_fit)[[1]][1,4] #from the anova table

fstar>crit.val
```


###6.
*In order to ensure the assumptions of the model from part 4 are valid diagnostics must be conducted. These diagnostics a histogram of the reisudlas and studentized residuals, a QQ Plot, a scatter plot of the residuals and Bartlett's test for homogeneity*


*The scatter plot of the residuals show that the residuals have an unequal variance. Moreover, the QQ Plot shows some signgicant deviations from normality and the histogram of the residuals and the studentized residuals show a heavy right tail. Furthmore, the Bartlett test of homogeneity of variances was conducted and the pvaule of the test was less than 2.2e-16 which is less than the signficance level 0.05. Meaning there is  evidence of significantly unequal variances of error terms. Therefore our assumption of normality and equal variance does not hold.*


```{r, echo = FALSE}
par(mfrow=c(1,3))
# obaain the residulas for the model from pt 4
residuals=anova_fit$residuals
hist(residuals)

#calculate the MSE
mse=sum(anova_fit$residuals^2)/anova_fit$df.residual




# Semistudentized residuals

residuals_semistd=anova_fit$residuals/sqrt(mse);
hist(residuals_semistd)

#calculate studentized residuals
student_resids <- studres(anova_fit)
hist(student_resids)


```


```{r, echo = FALSE}


par(mfrow=c(1,2))
qqnorm(residuals)
qqline(residuals)
plot(residuals~anova_fit$fitted.values,type='p',pch=16,cex=1.5,xlab="Fitted values",ylab="Residuals")
abline(0,0)


```



#####Bartlett Test Results:
```{r, echo=F}
#preform bartlett test to check for equal var
bartlett.test(wage ~ occupation, data = Wage)

```


  
###7.
Using Tukey's Honestly Significant Difference CI Test it is evident that the managment occupation has the higest mean wage. Looking at all the intervals that compare managment with the other occupations almost all, with the expection of the technical-management comparison, a have an adjusted pvaule of less than 0.05.
```{r, echo = FALSE}
alpha = 0.05
tukey_CI=TukeyHSD(anova_fit,conf.level = 1-alpha)

knitr::kable(data.frame(tukey_CI$occupation))

par(mar=c(7,9,5,4))
plot(tukey_CI, las=2 , col="red",pch=9, font = 1,cex.axis=.8)

#tukey_CI$occupation['technical-management',]
```

*Table that shows that the occupation with the highest mean wage is managemet*
```{r, echo = FALSE}

means_comb=tapply( Wage$wage,Wage$occupation,mean)

knitr::kable((means_comb), col.names = c("Mean Wage"))


```



###8.

*The Loss function would be equal to SSE (Sum of Squares Error):*
$$ L_1(\mu,\alpha_i)  = \sum_{i=1}^{r} \sum_{j=1}^{n_i} (Y_{ij} - \alpha_i - \mu)^2 \quad for\space i=1,..,6$$



*The Likelihood function:*
$$l_2(\mu,\alpha_i)  = \prod_{i=1}^r \prod_{j=1}^{n_i} \frac{1}{\sqrt{2\pi \sigma^2} } \exp\left[-\frac{ (Y_{i,j} - \mu-\alpha_i)^2}{2\sigma^2}\right]\quad for\space i=1,..,6$$
*Taking the log of $l_2(\mu,\alpha)$ will result in the Log Likelihood function.*

*The Log likelihood:*
$$L_2(\mu,\alpha) = \sum_{i=1}^r \sum_{j=1}^{n_i} \log\left\{ \frac{1}{\sqrt{2\pi \sigma^2} } \exp\left[-\frac{ (Y_{i,j} - \mu-\alpha_i)^2}{2\sigma^2} \right]  \right\}$$

```{r, echo = FALSE}

```



###9.


$$  (\hat\mu, \hat\alpha)=   argmax_{\mu,\alpha}L_2(\mu,\alpha)\quad s.t. \sum_{i=1}^r n_i\alpha_i=0  $$





*Start by breaking down the log likelihood funtion to*

$$ L_2(\mu,\alpha) =  -n\log (\frac{1}{\sqrt{2\pi \sigma^2}})  - \frac{ \sum_{i=1}^r \sum_{j=1}^{n_i}(Y_{i,j} - \mu-\alpha_i)^2}{2\sigma^2} $$



*To maximize the parameters the partial dervative of the log-likelihood is set equal to 0*
$$\frac{\partial L_2}{\partial \alpha} L_2(\mu,\alpha)  = 0 $$

$$\frac{\partial L_2}{\partial  \mu} L_2(\mu,\alpha)  = 0 $$

Thus the MLE is the lest square estimators from question 2:
The MLE for $\hat\alpha$
$$ \hat\alpha =  \hat\mu_i - \hat\mu \Rightarrow \hat\alpha= \bar{Y}_{i .}-\bar{Y}_{. .}$$

The MLE for $\hat\mu$

$$\frac{\partial L_2}{\partial  \mu} L_2(\mu,\alpha) =\frac{\sum_{i=1}^r \sum_{j=1}^{n_i}(Y_{i,j}- \alpha_i -\mu) }{\sigma}= 0 $$



$$  \hat\mu = \sum_{i=1}^r \sum_{j=1}^{n_i}(Y_{i,j}- \alpha_i)$$


##Acknowledgement
###Session information
```{r}
sessionInfo()

```


#####Appendix of Code:
```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```
